### 2011-06-05 docs/pubs/EIM-2011/abstract.tex

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechansism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

>>>>

\begin{abstract}
  Effective discovery and integration of ecological data within data
  management systems requires rich semantic information that can
  describe and relate the types of information contained within
  disparate data sets. Within the Semtools project, we have developed
  approaches for expressing and representing semantic annotations of
  data sets for supplementing attribute and data-level metadata with
  terms drawn from domain-specific ontologies. Annotations provide a
  formal mechanism that can be used together with reasoning systems
  to enhance existing data discovery and integration approaches.  In
  this paper we describe extensions to the Ecological Metadata
  Language (EML) and associated tools for storing and using semantic
  annotations.  Specifically, we describe new user interface
  components implemented within the Morpho metadata editor for
  capturing user-supplied semantic annotations, extensions to the
  Metacat system for storing and accessing annotations and
  corresponding OWL-DL ontologies, and a new API within Metacat that
  uses annotation metadata to provide concept-based search and
  integration of data sets.
%   The Data Manager paper from the last EIM included semantic
%   foreshadowing in the 'future directions' section and this is a
%   logical continuation of that work. Previously, we saw that the DM
%   library enabled data integration tasks provided we knew exactly
%   which data attributes from exactly which datasets were "compatible"
%   with each other. Here "compatible" means not simply structurally
%   similar (e.g. decimal numbers) but also conceptually similar
%   (e.g. ocean water temperature in C). By including rich semantic
%   annotations we initially show simple integration capabilities where
%   the subject domain of our data corpus is relatively constrained
%   (i.e. SBC-LTER) and the observational models represented are
%   minimally variable (i.e. context excluded). Semantically-enahnced
%   query capabilities facilitate this 'smart integration' in multiple
%   ways.
%   First, the corpus is refined through compound concept-based
%   queries that exercise OBOE-compatible ontological subsumption
%   hierarchies to locate roughly equivalent attributes across
%   datasets. Additional query constraints are defined in relation to
%   the precise observational model formally expressed in OBOE. Finally,
%   the actual data values are interrogated so that only those that fall
%   in the desired range are returned. From this resultant subset of
%   disparate data objects we create a single synthetic data product
%   that represents a 'materialization' of the observations. Technical
%   discussion will include high-level descriptions of the
%   infrastructure and tools employed with focus on our development in
%   the rapidly evolving semantic frontier where scalability and best
%   practices are particularly germane and equally elusive.
\end{abstract}

### 2011-06-05 docs/pubs/EIM-2011/application.tex

\mypara{Concept query.} The semantic query interface (see
\figref{fig:metacat-query}) is implemented as a Web application over
Metacat that primarily supports locating datasets by how well their
observational models match the given criteria. The interface provides
\emph{structured} as opposed to \emph{unstructured}, i.e.,
keyword-based queries. In particular, query criteria given by users
largely mirror the structure of a semantic annotation in that
combinations of Entity, Characteristic, and Protocol are specified and
optionally compounded when increased precision is sought.

@@

Using compound semantic query criteria applies a finer-grained filter
on the datasets that are returned. Results can be restricted to only
those datasets that include measurements for a set of specific
characteristics of a particular observational entity. Furthermore, a
query can specify that those measurements come from precisely the
\emph{same instance of that entity}; a feature that fully exercises
the comprensive observational structure expressed in the annotation
and enables higher query precision as described above.

\mypara{Data-level query.} For even more precise recall, the OBOE
model can be (partially) materialized (see above) during the query
stage after which a data range filter can be applied. Different
techniques are available for merging the annotation with the data that
it describes, but for performance reasons a hybrid approach has been
adopted in which premliminary search results from a structured query
are collated and only that subset is materialized. Because our corpus
is described using EML in conjunction with the annotation syntax, the
Data Manager Library \cite{leinfelder10} is used to load the described
raw data (into a relational database) while the annotation informs the
correct use of the Data Manager query and filtering features. For any
measurements that match the concept query criteria, we verify that
those measurements (e.g., attributes) contain data values within the
range specifed in the initial semantic data query and return the data
packages that contain them (see \figref{fig:metacat-query}).

\mypara{Data integration.} The materialization routine used for
semantic data queries laid the groundwork for enabling data
integration. In addition to inspecting the data for values within a
range and returning the datasets that contain a match, the data
integration feature of the Web application goes further by
constructing a synthetic data product (table) that represents the
complete results of the query in terms of both the attributes and the
values that are returned. Each original data set may have very
different syntactic structures (e.g. column number, naming, order) but
could share a subset of attributes that are semantically compatible as
defined in accompanying annotations. These compatible, in-common
attributes become the facets and metrics in a synthetic dataset; their
values filtered accordingly. \figref{fig:integration} further
illustrates the data integration support provided in the current
implementation of the Web application. Consider two data packages,
denoted by A and B in \figref{fig:integration}.  Annotations (denoted
C and D) are used to determine semantically equivalent data attributes
contained in the data sets (denoted by E and F). The attributes {\tt
  plot} and {\tt site} are considered equivalent measurements of the
characteristic Location; attributes {\tt weight} and {\tt wt} both map
to the same characteristic Mass. The Semantic Mediation API parses
each annotation and then computes an equivalance mapping among
attributes based on their corresponding annotations. The Data Manager
Library is then used to load the data sets and then query each data
set to produce and merge the final, synthetic result data set.

>>>>

\mypara{Concept query.} The semantic query interface (see
\figref{fig:metacat-query}) is implemented as a Web application over
Metacat that primarily supports locating data sets by how well their
observational models match the given criteria. The interface provides
\emph{structured} as opposed to \emph{unstructured}, i.e.,
keyword-based queries. In particular, query criteria given by users
largely mirror the structure of a semantic annotation in that
combinations of Entity, Characteristic, and Protocol are specified and
optionally compounded when increased precision is sought.

@@

Using compound semantic query criteria applies a finer-grained filter
on the data sets that are returned. Results can be restricted to only
those data sets that include measurements for a set of specific
characteristics of a particular observational entity. Furthermore, a
query can specify that those measurements come from precisely the
\emph{same instance of that entity}; a feature that fully exercises
the comprehensive observational structure expressed in the annotation
and enables higher query precision as described above.

\mypara{Data-level query.} For even more precise recall, the OBOE
model can be (partially) materialized (see above) during the query
stage after which a data range filter can be applied. Different
techniques are available for merging the annotation with the data that
it describes, but for performance reasons a hybrid approach has been
adopted in which preliminary search results from a structured query
are collated and only that subset is materialized. Because our corpus
is described using EML in conjunction with the annotation syntax, the
Data Manager Library \cite{leinfelder10} is used to load the described
raw data (into a relational database) while the annotation informs the
correct use of the Data Manager query and filtering features. For any
measurements that match the concept query criteria, we verify that
those measurements (e.g., attributes) contain data values within the
range specified in the initial semantic data query and return the data
packages that contain them (see \figref{fig:metacat-query}).

\mypara{Data integration.} The materialization routine used for
semantic data queries laid the groundwork for enabling data
integration. In addition to inspecting the data for values within a
range and returning the data sets that contain a match, the data
integration feature of the Web application goes further by
constructing a synthetic data product (table) that represents the
complete results of the query in terms of both the attributes and the
values that are returned. Each original data set may have very
different syntactic structures (e.g. column number, naming, order) but
could share a subset of attributes that are semantically compatible as
defined in accompanying annotations. These compatible, in-common
attributes become the facets and metrics in a synthetic data set; their
values filtered accordingly. \figref{fig:integration} further
illustrates the data integration support provided in the current
implementation of the Web application. Consider two data packages,
denoted by A and B in \figref{fig:integration}.  Annotations (denoted
C and D) are used to determine semantically equivalent data attributes
contained in the data sets (denoted by E and F). The attributes {\tt
  plot} and {\tt site} are considered equivalent measurements of the
characteristic Location; attributes {\tt weight} and {\tt wt} both map
to the same characteristic Mass. The Semantic Mediation API parses
each annotation and then computes an equivalence mapping among
attributes based on their corresponding annotations. The Data Manager
Library is then used to load the data sets and then query each data
set to produce and merge the final, synthetic result data set.

### 2011-06-05 docs/pubs/EIM-2011/framework.tex

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
\caption{Partial OBOE semantic annotation for Kelp sampling
  data. Shaded nodes represent ontological concepts; rectangular nodes
  are data table attibutes mapped to OBOE measurement
  characteristics.}
\label{fig:kelp-mass-model}
\end{figure}

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows users to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierachical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated and fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., obserations). The editor
provides a number of additional features including allowing users to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

\mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
augements Metacat's existing metadata storage and search by allowing
annotations to be saved and queried alongside the metadata and data
that they annotate. In addition to traditional keyword and spatial
search criteria, the Metacat plugin allows semantic criteria to be
included where they may either increase query recall using
term-expansion (i.e., traversing the class subsumption hierarchy) or
refine the resultset by limiting matches to datasets that contain the
specified observational model (e.g., combinations of OBOE-compatible
entity, characteristic, measurement standard, or protocol
concepts). The observational model can be leveraged further by
materializing the annotation and data artifact (via the Data Manager
Library \cite{leinfelder10:_metad_driven_approac_to_loadin}) into a
fully instantiated OBOE model and inspecting (and querying over) the
observational values themselves.

>>>>

\begin{figure}[!b]
\centering
\includegraphics[width=0.5\textwidth]{images/kelp-mass-model.png}
\caption{Partial OBOE semantic annotation for Kelp sampling
  data. Shaded nodes represent ontological concepts; rectangular nodes
  are data table attributes mapped to OBOE measurement
  characteristics.}
\label{fig:kelp-mass-model}
\end{figure}

@@

\mypara{The Morpho Editor Plugin.}  The semantic-annotation editor
plugin for Morpho provides a front-end to the Semantic Mediation API
and allows users to define annotations for existing data package
descriptions. The editor provides a simple ``fill-in-the-blank'' style
form-based interface with a searchable hierarchical concept selection
widget (see \figref{fig:morpho-annotation}). The plugin seamlessly
integrates with a standard Morpho installation and provides semantic
query capabilities for locating data packages, marking up data
packages with semantic annotations, and saving annotations locally or
to a shared repository where they can be discovered and explored by
other users. The annotation editor in Morpho allows a user to view the
data set being annotated and fill in (by selecting an appropriate
ontology term) the characteristic, measurement standard, protocol, and
associated entity for each data-set attribute. Users can also specify
whether an observation spans multiple columns, and can provide context
relationships between attributes (i.e., observations). The editor
provides a number of additional features including allowing users to
view the entire annotation (similar to \figref{fig:kelp-mass-model})
and to specify additional mapping constraints for observations and
measurements.

\mypara{Metacat Query Extensions.}  The semantic plugin for Metacat
augments Metacat's existing metadata storage and search by allowing
annotations to be saved and queried alongside the metadata and data
that they annotate. In addition to traditional keyword and spatial
search criteria, the Metacat plugin allows semantic criteria to be
included where they may either increase query recall using
term-expansion (i.e., traversing the class subsumption hierarchy) or
refine the result set by limiting matches to data sets that contain
the specified observational model (e.g., combinations of
OBOE-compatible entity, characteristic, measurement standard, or
protocol concepts). The observational model can be leveraged further
by materializing the annotation and data artifact (via the Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin})
into a fully instantiated OBOE model and inspecting (and querying
over) the observational values themselves.

### 2011-06-05 docs/pubs/EIM-2011/intro.tex

A major challenge in environmental information management concerns
providing effective approaches for the discovery and integration of
heteregenous data sets. For instance, locating and combining relevant
observational data are often critical and time-consuming steps for
researchers studying phenomena at broad spatial, temporal, and
biological scales
\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do,green05:_compl_in_ecolog_and_conser,sorokina09:_detec_inter_variab_inter_obser_ornit_data,jackson01:_histor_overf_and_recen_collap}. The
underlying data sets used within such studies frequently differ in
subtle and complex ways, due in part to the protocols used for data
collection, the types of observations made, and the experimental and
other contextual information associated with the data set. These
differences in turn can lead to structural and semantic heterogeneity
among data sets that make them hard to discover using current data
management approaches and require considerable manual effort by
researchers needing to combine data sets.

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying datasets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often impede integration across scientific data sets.

In this paper we describe extensions to the Ecological Metadata
Language (EML) \cite{Fegraus07} and suppporting tools for enabling
improved discovery and integration of ecological data sets. Our work
is based on the Extensible Observations Ontology (OBOE)
\cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
generic observational model implemented in OWL-DL for describing
domain-specific observation and measurement types. Our approach adds
additional metadata in the form of semantic annotations that link
attributes within datasets to OBOE terms for describing the implicit
observation and measurement types found within data sets. Semantic
annotations are executable in the sense that they can be used to
convert a data set into a collection of observation and measurement
instances, providing a more uniform representation for expressing
queries and performing integration.  To support the creation of
annotations, we have extended the Morpho metadata editor
\cite{metacat02:_manag_heter_ecolog_data_using_morph} with a
high-level user interface as well as the Metacat data catalog
\cite{berkley01:_metac} for storing and querying annotations through a
new Semantic Mediation API. This API can also be used to perform basic
data-level integration tasks using our prior work on the EML Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin}.

>>>>

A major challenge in environmental information management concerns
providing effective approaches for the discovery and integration of
heterogeneous data sets. For instance, locating and combining relevant
observational data are often critical and time-consuming steps for
researchers studying phenomena at broad spatial, temporal, and
biological scales
\cite{worm06:_impac_of_biodiv_loss_ocean_ecosy_servic,pennings05:_do,green05:_compl_in_ecolog_and_conser,sorokina09:_detec_inter_variab_inter_obser_ornit_data,jackson01:_histor_overf_and_recen_collap}. The
underlying data sets used within such studies frequently differ in
subtle and complex ways, due in part to the protocols used for data
collection, the types of observations made, and the experimental and
other contextual information associated with the data set. These
differences in turn can lead to structural and semantic heterogeneity
among data sets that make them hard to discover using current data
management approaches and require considerable manual effort by
researchers needing to combine data sets.

A number of recent efforts within the earth and environmental
informatics communities are adopting the notion of an \emph{observation} as a
key modeling concept for enabling improved discovery and integration
of scientific data
\cite{om,fox09:_ontol,tarboton07:_cuahs_commun_obser_data_model,cushing07:_compon_based_end_user_datab,balhoff10:_phenex,bowers08}. These
approaches provide higher-level observational data models for
describing and representing observations and measurements found in
underlying data sets by defining common ``core'' concepts such as the
entities or features being observed, measurement units and protocols,
and context relationships between observations \cite{om,bowers08}.  A
major goal of these approaches is to enable interoperability and
uniform access to data by abstracting away the underlying
representation details that often impede integration across scientific data sets.

In this paper we describe extensions to the Ecological Metadata
Language (EML) \cite{Fegraus07} and supporting tools for enabling
improved discovery and integration of ecological data sets. Our work
is based on the Extensible Observations Ontology (OBOE)
\cite{bowers08,madin07:_ontol_for_descr_and_synth}, which represents a
generic observational model implemented in OWL-DL for describing
domain-specific observation and measurement types. Our approach adds
additional metadata in the form of semantic annotations that link
attributes within data sets to OBOE terms for describing the implicit
observation and measurement types found within data sets. Semantic
annotations are executable in the sense that they can be used to
convert a data set into a collection of observation and measurement
instances, providing a more uniform representation for expressing
queries and performing integration.  To support the creation of
annotations, we have extended the Morpho metadata editor
\cite{metacat02:_manag_heter_ecolog_data_using_morph} with a
high-level user interface as well as the Metacat data catalog
\cite{berkley01:_metac} for storing and querying annotations through a
new Semantic Mediation API. This API can also be used to perform basic
data-level integration tasks using our prior work on the EML Data
Manager Library \cite{leinfelder10:_metad_driven_approac_to_loadin}.

### 2011-06-05 docs/pubs/EIM-2011/summary.tex

The Semtools project has been successful in exploring and codifying technologies and techniques for applying semantic concepts to observational data. By providing mechanisms for linking datasets to ontological terms organized in a high-level observational model (e.g., OBOE), these new extensions to Metacat and Morpho help to overcome a number of limitations in
existing metadata management systems that strive to provide effective data discovery and integration features. Our close involvement with the SONet Project (Scientific Observations Network) \cite{sonet} encourages continued use-case refinement that will inform future semantic tool development and place an emphasis on intuitive interfaces and incremental adoption. This varied community of stakeholders is firmly invested in the use of cutting edge semantic solutions that will ultimately benefit multiple science disciplines by reducing obstacles to broad data sharing and innovative reuse.

>>>>

The Semtools project has been successful in exploring and codifying
technologies and techniques for applying semantic concepts to
observational data. By providing mechanisms for linking data sets to
ontological terms organized in a high-level observational model (e.g.,
OBOE), these new extensions to Metacat and Morpho help to overcome a
number of limitations in existing metadata management systems that
strive to provide effective data discovery and integration
features. Our close involvement with the SONet Project (Scientific
Observations Network) \cite{sonet} encourages continued use-case
refinement that will inform future semantic tool development and place
an emphasis on intuitive interfaces and incremental adoption. This
varied community of stakeholders is firmly invested in the use of
cutting edge semantic solutions that will ultimately benefit multiple
science disciplines by reducing obstacles to broad data sharing and
innovative reuse.
